---
title: "Model_Definitions"
output:
  html_document:
    df_print: paged
---

This is code to define all the different POMP models to consider.

```{r forecast, echo=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(pomp))
suppressPackageStartupMessages(library(ggplot2))
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Read In Data - Process Opponent Strength

```{r}
all_games <- readxl::read_xlsx("mlb-2024-asplayed.xlsx", sheet = "As-Played Schedule")
all_games$Away_Starter <- all_games$`Away Starter`
all_games$Home_Starter <- all_games$`Home Starter`
all_games$Away_Score <- all_games$`Away Score`
all_games$Home_Score <- all_games$`Home Score`
det_games <- all_games[all_games$Away=="Detroit Tigers"|all_games$Home=="Detroit Tigers", ]
det_games$opp_strength <- NA
for (row in rownames(det_games)) {
  game <- det_games[row,]
  if (game$Away=="Detroit Tigers") {
    opp <- game$Home
    opp_p <- game$Home_Starter
  } else {
    opp <- game$Away
    opp_p <- game$Away_Starter
  }
  opp_pitch_games <- rbind(
    all_games %>% filter(Away==opp,
                         Home!="Detroit Tigers",
                         Away_Starter==opp_p) %>%
      mutate(RA=Home_Score),
    all_games %>% filter(Home==opp,
                         Away!="Detroit Tigers",
                         Home_Starter==opp_p) %>%
      mutate(RA=Away_Score)
  )
  if (nrow(opp_pitch_games>0)) {
    det_games[row, "opp_strength"] <- mean(opp_pitch_games$RA)
  } else {
    opp_games <- rbind(
      all_games %>% filter(Away==opp,
                           Home!="Detroit Tigers") %>%
        mutate(RA=Home_Score),
      all_games %>% filter(Home==opp,
                           Away!="Detroit Tigers") %>%
        mutate(RA=Away_Score)
    )
    det_games[row, "opp_strength"] <- mean(opp_games$RA)
  }
}

det_games$R <- ifelse(det_games$Home=="Detroit Tigers",
                      det_games$Home_Score, det_games$Away_Score)
det_runs <- data.frame(game=(1:162), R=det_games$R)

```










# Get Initial Dispersion Parameter - Neg Binom

```{r}
mean_2023 <- 661.0/162.0
var_2023 <- 7.20470056

disp_2023 <- (mean_2023^2)/(var_2023-mean_2023)
round(disp_2023,2)
```




```{r}
run_level <- "explore"
if (run_level=="explore") {
  nseq <- 5
  length <- 10
  nprof <- 10
} else if (run_level=="final") {
  nseq <- 500
  length <- 40
  nprof <- 15
}
```



# Initialize POMP Object Components
```{r}

runs_step <- list(
  AR1 = Csnippet("
    double d_X = rnorm(0, sigma);
    X = phi*X + d_X;
  "),
  static = Csnippet("
    // X stays constant
  ")
)

runs_rinit <- list(
  AR1 = Csnippet("
    X = 0;
  "),
  static = function(...) {
    setNames(numeric(0), character(0))
  }
)



runs_dmeas <- list(
  AR1_pois = Csnippet("
    double lambda;
    lambda = exp(X + mu + gamma*(opp_strength-4.6));
    lik = dpois(R, lambda, give_log);
  "),
  static_pois = Csnippet("
    double lambda;
    lambda = exp(mu + gamma*(opp_strength-4.6));
    lik = dpois(R, lambda, give_log);
  "),
  AR1_nbin = Csnippet("
    double lambda;
    lambda = exp(X + mu + gamma*(opp_strength-4.6));
    lik = dnbinom_mu(R, k, lambda, give_log);
  "),
  static_nbin = Csnippet("
    double lambda;
    lambda = exp(mu + gamma*(opp_strength-4.6));
    lik = dnbinom_mu(R, k, lambda, give_log);
  ")
)



runs_rmeas <- list(
  AR1_pois = Csnippet("
    double lambda;
    lambda = exp(X + mu + gamma*(opp_strength-4.6));
    R = rpois(lambda);
  "),
  static_pois = Csnippet("
    double lambda;
    lambda = exp(mu + gamma*(opp_strength-4.6));
    R = rpois(lambda);
  "),
  AR1_nbin = Csnippet("
    double lambda;
    lambda = exp(X + mu + gamma*(opp_strength-4.6));
    R = rnbinom_mu(k, lambda);
  "),
  static_nbin = Csnippet("
    double lambda;
    lambda = exp(mu + gamma*(opp_strength-4.6));
    R = rnbinom_mu(k, lambda);
  ")
)

partrans <- parameter_trans(
  log = "sigma"  # Constrain sigma to be > 0
)

# Create Covariate Table
covar <- covariate_table(
  time = c(0, det_runs$game),
  opp_strength = c(0, det_games$opp_strength),
  times = "time"
)
```
#HERE
Then we fold the data and model components into a `pomp` object.
```{r fold_it}
set.seed(657)
runsPOMP <- NULL

det_runs |>
  pomp(times="game",t0=0,
       rprocess=euler(runs_step[["AR1"]],delta.t=1),
       rinit=runs_rinit[["AR1"]],
       rmeasure=runs_rmeas[["AR1_pois"]],
       dmeasure=runs_dmeas[["AR1_pois"]],
       covar = covar,
       statenames="X",
       paramnames=c("gamma","phi","sigma", "mu"),
       obsnames = "R",
       partrans = partrans
  ) -> runsPOMP[["AR1_pois"]]
coef(runsPOMP[["AR1_pois"]]) <- c(gamma=-.25, phi=1, sigma=0.005, mu=log(661.0/162.0))

det_runs |>
  pomp(times="game",t0=0,
       rprocess=euler(runs_step[["static"]],delta.t=1),
       rinit=runs_rinit[["static"]],
       rmeasure=runs_rmeas[["static_pois"]],
       dmeasure=runs_dmeas[["static_pois"]],
       covar = covar,
       paramnames=c("gamma","mu"),
       obsnames = "R"
  ) -> runsPOMP[["static_pois"]]
coef(runsPOMP[["static_pois"]]) <- c(gamma=-1, mu=log(661.0/162.0))



det_runs |>
  pomp(times="game",t0=0,
       rprocess=euler(runs_step[["AR1"]],delta.t=1),
       rinit=runs_rinit[["AR1"]],
       rmeasure=runs_rmeas[["AR1_nbin"]],
       dmeasure=runs_dmeas[["AR1_nbin"]],
       covar = covar,
       statenames="X",
       paramnames=c("gamma","phi", "sigma", "k", "mu"),
       obsnames = "R",
       partrans = partrans
  ) -> runsPOMP[["AR1_nbin"]]
coef(runsPOMP[["AR1_nbin"]]) <- 
  c(gamma=-1, phi=1, sigma=0.1, k=disp_2023, mu=log(661.0/162.0))


det_runs |>
  pomp(times="game",t0=0,
       rprocess=euler(runs_step[["static"]],delta.t=1),
       rinit=runs_rinit[["static"]],
       rmeasure=runs_rmeas[["static_nbin"]],
       dmeasure=runs_dmeas[["static_nbin"]],
       covar = covar,
       paramnames=c("gamma", "k", "mu"),
       obsnames = "R"
  ) -> runsPOMP[["static_nbin"]]
coef(runsPOMP[["static_nbin"]]) <- c(gamma=-1, k=disp_2023, mu=log(661.0/162.0))
#coef(runsPOMP[["static_nbin"]]) <- c(gamma=-1, k=disp_2023)


```




```{r library_invoke, echo=FALSE}
suppressPackageStartupMessages(library(foreach))
suppressPackageStartupMessages(library(doFuture))
plan(multisession)
Output <- NULL
```

```{r}

global_guesses <- function(model) {
  if (model=="AR1_pois") {
    runif_design(
      lower=c(gamma=-1, phi=-0.25, sigma=0.0001, mu=0.001),
      upper=c(gamma=1, phi=1.5, sigma=0.6, mu=log(7)),
      nseq=nseq
    ) -> guesses
  } else if (model=="AR1_nbin") {
    runif_design(
      lower=c(gamma=-1,phi=-0.25,sigma=0.0001, k=1, mu=0.001),
      upper=c(gamma=1,phi=1.5,sigma=0.6, k=10, mu=log(7)),
      nseq=nseq
    ) -> guesses
  } else if (model=="static_pois") {
    runif_design(
      lower=c(gamma=-1, mu=0.001),
      upper=c(gamma=1, mu=log(7)),
      nseq=nseq
    ) -> guesses
  } else if (model=="static_nbin") {
    runif_design(
      lower=c(gamma=-1, k=1, mu=0.001),
      upper=c(gamma=1, k=10, mu=log(7)),
      nseq=nseq
    ) -> guesses
  }
  return(guesses)
}

rw_trans_models <- function(model) {
  if (model=="AR1_pois") {
    rw_sd_model <- 
      rw_sd(gamma = 0.02, phi = 0.02, sigma = ivp(0.02), mu=ivp(0.02))
    trans_model <- parameter_trans(log = c("sigma", "mu"))
  } else if (model=="AR1_nbin") {
    rw_sd_model <- 
      rw_sd(gamma = 0.02, phi = 0.02, sigma = ivp(0.02), k = ivp(0.02), mu=ivp(0.02))
    trans_model <- parameter_trans(log = c("sigma", "k", "mu"))
  } else if (model=="static_pois") {
    rw_sd_model <- 
      rw_sd(gamma = 0.02, mu = ivp(0.02))
    trans_model <- parameter_trans(log = "mu")
  } else if (model=="static_nbin") {
    rw_sd_model <- 
      rw_sd(gamma = 0.02, k = ivp(0.02), mu=ivp(0.02))
    trans_model <- parameter_trans(log = c("k", "mu"))
  }
  return(list(rw_sd_model=rw_sd_model,
              trans_model=trans_model))
}


```


# AR1:Poisson Model
```{r AR1_Pois}
model <- "AR1_pois"
csv_name <- paste0("run_parms_", model, ".csv")
Output[[model]] <- NULL
```


```{r}
runsPOMP[[model]] |>
  simulate(
    params=coef(runsPOMP[[model]]),
    nsim=20,format="data.frame", include.data=TRUE
  ) -> sims_init

Output[[model]][["sims_init"]] <- sims_init

set.seed(657)
sims_init |>
  ggplot(aes(x=game,y=R,group=.id,color=.id=="data"))+
  geom_line()+
  guides(color="none")
```



# Plot ESS
```{r}
runsPOMP[[model]] |> 
  pfilter(Np=2000) -> pf
plot(pf)
min_ESS_init <- min(pf@eff.sample.size)
min_ESS_init

Output[[model]][["pf"]] <- pf
Output[[model]][["min_ESS_init"]] <- min_ESS_init
```

```{r}
set.seed(10403)
foreach(i=1:10,.combine=c, 
        .options.future=list(seed=TRUE) 
        ) %dofuture% {
          runsPOMP[[model]] |> pfilter(Np=5000) 
          } -> pf  
pf |> logLik() |> logmeanexp(se=TRUE) -> L_pf 
L_pf

Output[[model]][["L_pf"]] <- L_pf
```


```{r}
pf[[1]] |> coef() |> bind_rows() |> 
  bind_cols(loglik=L_pf[1],loglik.se=L_pf[2]) |> 
  write_csv(csv_name)
```

We'll now carry out a local search using `mif2` around this point in our parameter space.

```{r}
mif_sets <- rw_trans_models(model)

foreach(i=1:20,.combine=c, 
        .options.future=list(seed=3318) 
        ) %dofuture% { runsPOMP[[model]] |> 
            mif2( Np=2000, Nmif=150, 
                  cooling.fraction.50 = 0.5,
                  rw.sd = mif_sets$rw_sd_model,
                  partrans = mif_sets$trans_model,
                  paramnames = names(coef(runsPOMP[[model]]))
                  )  
          } -> mifs_local
Output[[model]][["mifs_local"]] <- mifs_local
```

```{r}
mifs_local |>
  traces() |>
  melt() |>
  ggplot(aes(x=iteration,y=value,group=.L1,color=factor(.L1)))+
  geom_line()+
  guides(color="none")+
  facet_wrap(~name,scales="free_y") +
  labs(x = "Iteration", y = "Value", title="Local Search Diagnostic Plots") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
foreach(mf=mifs_local,.combine=rbind,
    .options.future=list(seed=101)
  ) %dofuture% {
evals <- replicate(10, logLik(pfilter(mf,Np=5000)))
  ll <- logmeanexp(evals,se=TRUE)
  mf |> coef() |> bind_rows() |>
    bind_cols(loglik=ll[1],loglik.se=ll[2])
} -> results

#store in our dataframe
read_csv(csv_name, show_col_types = FALSE) |>
  bind_rows(results) |>
  arrange(-loglik) |>
  write_csv(csv_name)


Output[[model]][["results_loc"]] <- results
print(max(results$loglik))
```



```{r}
lik_form <- "~loglik"
for (p in names(coef(runsPOMP[[model]]))) {
  lik_form <- paste0(lik_form, "+", p)
}
lik_form <- as.formula(lik_form)

pairs(lik_form,data=results, pch=16)
```


```{r}
set.seed(657)
guesses <- global_guesses(model)
mf1 <- mifs_local[[1]]
```

```{r}
foreach(guess=iterators::iter(guesses,"row"), .combine=rbind,
    .options.future=list(seed=1312)
) %dofuture% {
    mf1 |>
      mif2(params=c(guess)) |>
      mif2(Nmif=100) -> mf
    replicate(
      10,
      mf |> pfilter(Np=5000) |> logLik()
  ) |>
    logmeanexp(se=TRUE) -> ll
  mf |> coef() |> bind_rows() |>
    bind_cols(loglik=ll[1],loglik.se=ll[2])
} -> results

Output[[model]][["results_glob"]] <- results
```



```{r}
print(max(results$loglik))
```



```{r}
results |>
  filter(loglik>max(loglik)-50) |>
  bind_rows(guesses) |>
  mutate(type=if_else(is.na(loglik),"guess","result")) |>
  arrange(type) -> all

Output[[model]][["all"]] <- all

pairs(lik_form, data=all, pch=16, cex=0.3,
      col=ifelse(all$type=="guess",grey(0.5),"red"))
```




## Profile Likelihood

```{r AR1_Pois_Prof}
results |>
  filter(loglik>max(loglik)-50) |>
  bind_rows(guesses) |>
  mutate(type=if_else(is.na(loglik),"guess","result")) |>
  arrange(type) -> all

all |>
  filter(type=="result") |>
  filter(loglik>max(loglik)-10) |>
  ggplot(aes(x=phi,y=loglik))+
  geom_point() +
  labs(
  x=expression(phi),
  y="Log Likelihood",
  title="Poor Man’s Profile Likelihood - Phi") + 
  theme(plot.title = element_text(hjust = 0.5))
```



```{r}
read_csv(csv_name, show_col_types = FALSE) |>
  filter(loglik>max(loglik)-20,loglik.se<2) |>
  sapply(range) -> box
freeze(seed=97,
  profile_design(
    phi=seq(-0.25,0.99, length=length),
    lower=box[1,c("gamma", "sigma", "mu")],
    upper=box[2,c("gamma", "sigma", "mu")],
    nprof=nprof, type="runif"
  )) -> guesses
plot(guesses)
```


```{r}
foreach(guess=iterators::iter(guesses,"row"), .combine=rbind,
  .options.future=list(seed=1524)
) %dofuture% {
  mf1 |>
    mif2(params=c(guess),
         rw.sd=rw_sd(gamma=0.02, sigma=ivp(0.02), mu=ivp(0.02))) |>
    mif2(Nmif=100) -> mf
  replicate(
    10,
    mf |> pfilter(Np=5000) |> logLik()) |>
    logmeanexp(se=TRUE) -> ll
  mf |> coef() |> bind_rows() |>
    bind_cols(loglik=ll[1],loglik.se=ll[2])
} -> results

Output[[model]][["results_phi_prof"]] <- results

read_csv(csv_name, show_col_types = FALSE) |>
  bind_rows(results) |>
  filter(is.finite(loglik)) |>
  arrange(-loglik) |>
  write_csv(csv_name)

```


```{r}
maxloglik <- max(results$loglik,na.rm=TRUE)
ci.cutoff <- maxloglik-0.5*qchisq(df=1,p=0.95)

results |>
  filter(is.finite(loglik)) |>
  group_by(round(phi,5)) |>
  filter(rank(-loglik)<3) |>
  ungroup() |>
  ggplot(aes(x=phi,y=loglik))+
    geom_point()+
    #geom_smooth(method="loess",span=0.25)+
    geom_hline(color="red",yintercept=ci.cutoff) + 
    labs(x=expression(phi),
         y="Profile Log Likelihood",
         title="Profile Likelihood - Phi") + 
    theme(plot.title = element_text(hjust = 0.5))
```

```{r}
saveRDS(Output[[model]], paste0("Output_", model, ".RDS"))
```




# Static:Poisson Model
```{r Static_Pois}
model <- "static_pois"
csv_name <- paste0("run_parms_", model, ".csv")
Output[[model]] <- NULL
```


```{r}
set.seed(10403)
foreach(i=1:10,.combine=c, 
        .options.future=list(seed=TRUE) 
        ) %dofuture% {
          runsPOMP[[model]] |> pfilter(Np=5000) 
          } -> pf  
pf |> logLik() |> logmeanexp(se=TRUE) -> L_pf 
L_pf

Output[[model]][["L_pf"]] <- L_pf
```


```{r}
pf[[1]] |> coef() |> bind_rows() |> 
  bind_cols(loglik=L_pf[1],loglik.se=L_pf[2]) |> 
  write_csv(csv_name)
```

We'll now carry out a local search using `mif2` around this point in our parameter space.

```{r}
mif_sets <- rw_trans_models(model)

foreach(i=1:20,.combine=c, 
        .options.future=list(seed=3318) 
        ) %dofuture% { runsPOMP[[model]] |> 
            mif2( Np=2000, Nmif=50, 
                  cooling.fraction.50 = 0.5,
                  rw.sd = mif_sets$rw_sd_model,
                  partrans = mif_sets$trans_model,
                  paramnames = names(coef(runsPOMP[[model]]))
                  )  
          } -> mifs_local
Output[[model]][["mifs_local"]] <- mifs_local
```

```{r}
mifs_local |>
  traces() |>
  melt() |>
  ggplot(aes(x=iteration,y=value,group=.L1,color=factor(.L1)))+
  geom_line()+
  guides(color="none")+
  facet_wrap(~name,scales="free_y") +
  labs(x = "Iteration", y = "Value", title="Local Search Diagnostic Plots") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
foreach(mf=mifs_local,.combine=rbind,
    .options.future=list(seed=101)
  ) %dofuture% {
evals <- replicate(10, logLik(pfilter(mf,Np=5000)))
  ll <- logmeanexp(evals,se=TRUE)
  mf |> coef() |> bind_rows() |>
    bind_cols(loglik=ll[1],loglik.se=ll[2])
} -> results

#store in our dataframe
read_csv(csv_name, show_col_types = FALSE) |>
  bind_rows(results) |>
  arrange(-loglik) |>
  write_csv(csv_name)


Output[[model]][["results_loc"]] <- results
print(max(results$loglik))
```



```{r}
lik_form <- "~loglik"
for (p in names(coef(runsPOMP[[model]]))) {
  lik_form <- paste0(lik_form, "+", p)
}
lik_form <- as.formula(lik_form)

pairs(lik_form,data=results, pch=16)
```


```{r}
set.seed(657)
guesses <- global_guesses(model)
mf1 <- mifs_local[[1]]
```

```{r}
foreach(guess=iterators::iter(guesses,"row"), .combine=rbind,
    .options.future=list(seed=1312)
) %dofuture% {
    mf1 |>
      mif2(params=c(guess)) |>
      mif2(Nmif=100) -> mf
    replicate(
      10,
      mf |> pfilter(Np=5000) |> logLik()
  ) |>
    logmeanexp(se=TRUE) -> ll
  mf |> coef() |> bind_rows() |>
    bind_cols(loglik=ll[1],loglik.se=ll[2])
} -> results

Output[[model]][["results_glob"]] <- results
```



```{r}
print(max(results$loglik))
```



```{r}
results |>
  filter(loglik>max(loglik)-50) |>
  bind_rows(guesses) |>
  mutate(type=if_else(is.na(loglik),"guess","result")) |>
  arrange(type) -> all

Output[[model]][["all"]] <- all

pairs(lik_form, data=all, pch=16, cex=0.3,
      col=ifelse(all$type=="guess",grey(0.5),"red"))
```


```{r}
saveRDS(Output[[model]], paste0("Output_", model, ".RDS"))
```



#HERE
# AR1:NegBinom Model
```{r AR1:NBin}
model <- "AR1_nbin"
csv_name <- paste0("run_parms_", model, ".csv")
Output[[model]] <- NULL
```



```{r}
set.seed(10403)
foreach(i=1:10,.combine=c, 
        .options.future=list(seed=TRUE) 
        ) %dofuture% {
          runsPOMP[[model]] |> pfilter(Np=5000) 
          } -> pf  
pf |> logLik() |> logmeanexp(se=TRUE) -> L_pf 
L_pf

Output[[model]][["L_pf"]] <- L_pf
```


```{r}
pf[[1]] |> coef() |> bind_rows() |> 
  bind_cols(loglik=L_pf[1],loglik.se=L_pf[2]) |> 
  write_csv(csv_name)
```

We'll now carry out a local search using `mif2` around this point in our parameter space.

```{r}
mif_sets <- rw_trans_models(model)

foreach(i=1:20,.combine=c, 
        .options.future=list(seed=3318) 
        ) %dofuture% { runsPOMP[[model]] |> 
            mif2( Np=2000, Nmif=150, 
                  cooling.fraction.50 = 0.5,
                  rw.sd = mif_sets$rw_sd_model,
                  partrans = mif_sets$trans_model,
                  paramnames = names(coef(runsPOMP[[model]]))
                  )  
          } -> mifs_local
Output[[model]][["mifs_local"]] <- mifs_local
```

```{r}
mifs_local |>
  traces() |>
  melt() |>
  ggplot(aes(x=iteration,y=value,group=.L1,color=factor(.L1)))+
  geom_line()+
  guides(color="none")+
  facet_wrap(~name,scales="free_y") +
  labs(x = "Iteration", y = "Value", title="Local Search Diagnostic Plots") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
foreach(mf=mifs_local,.combine=rbind,
    .options.future=list(seed=101)
  ) %dofuture% {
evals <- replicate(10, logLik(pfilter(mf,Np=5000)))
  ll <- logmeanexp(evals,se=TRUE)
  mf |> coef() |> bind_rows() |>
    bind_cols(loglik=ll[1],loglik.se=ll[2])
} -> results

#store in our dataframe
read_csv(csv_name, show_col_types = FALSE) |>
  bind_rows(results) |>
  arrange(-loglik) |>
  write_csv(csv_name)


Output[[model]][["results_loc"]] <- results
print(max(results$loglik))
```



```{r}
lik_form <- "~loglik"
for (p in names(coef(runsPOMP[[model]]))) {
  lik_form <- paste0(lik_form, "+", p)
}
lik_form <- as.formula(lik_form)

pairs(lik_form,data=results, pch=16)
```


```{r}
set.seed(657)
guesses <- global_guesses(model)
mf1 <- mifs_local[[1]]
```

```{r}
foreach(guess=iterators::iter(guesses,"row"), .combine=rbind,
    .options.future=list(seed=1312)
) %dofuture% {
    mf1 |>
      mif2(params=c(guess)) |>
      mif2(Nmif=100) -> mf
    replicate(
      10,
      mf |> pfilter(Np=5000) |> logLik()
  ) |>
    logmeanexp(se=TRUE) -> ll
  mf |> coef() |> bind_rows() |>
    bind_cols(loglik=ll[1],loglik.se=ll[2])
} -> results

Output[[model]][["results_glob"]] <- results
```



```{r}
print(max(results$loglik))
```



```{r}
results |>
  filter(loglik>max(loglik)-50) |>
  bind_rows(guesses) |>
  mutate(type=if_else(is.na(loglik),"guess","result")) |>
  arrange(type) -> all

Output[[model]][["all"]] <- all

pairs(lik_form, data=all, pch=16, cex=0.3,
      col=ifelse(all$type=="guess",grey(0.5),"red"))
```


```{r}
saveRDS(Output[[model]], paste0("Output_", model, ".RDS"))
```





# Static:NegBinom Model
```{r Static_NBin}
model <- "static_nbin"
csv_name <- paste0("run_parms_", model, ".csv")
Output[[model]] <- NULL
```



```{r}
set.seed(10403)
foreach(i=1:10,.combine=c, 
        .options.future=list(seed=TRUE) 
        ) %dofuture% {
          runsPOMP[[model]] |> pfilter(Np=5000) 
          } -> pf  
pf |> logLik() |> logmeanexp(se=TRUE) -> L_pf 
L_pf

Output[[model]][["L_pf"]] <- L_pf
```


```{r}
pf[[1]] |> coef() |> bind_rows() |> 
  bind_cols(loglik=L_pf[1],loglik.se=L_pf[2]) |> 
  write_csv(csv_name)
```

We'll now carry out a local search using `mif2` around this point in our parameter space.

```{r}
mif_sets <- rw_trans_models(model)

foreach(i=1:20,.combine=c, 
        .options.future=list(seed=3318) 
        ) %dofuture% { runsPOMP[[model]] |> 
            mif2( Np=2000, Nmif=50, 
                  cooling.fraction.50 = 0.5,
                  rw.sd = mif_sets$rw_sd_model,
                  partrans = mif_sets$trans_model,
                  paramnames = names(coef(runsPOMP[[model]]))
                  )  
          } -> mifs_local
Output[[model]][["mifs_local"]] <- mifs_local
```

```{r}
mifs_local |>
  traces() |>
  melt() |>
  ggplot(aes(x=iteration,y=value,group=.L1,color=factor(.L1)))+
  geom_line()+
  guides(color="none")+
  facet_wrap(~name,scales="free_y") +
  labs(x = "Iteration", y = "Value", title="Local Search Diagnostic Plots") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
foreach(mf=mifs_local,.combine=rbind,
    .options.future=list(seed=101)
  ) %dofuture% {
evals <- replicate(10, logLik(pfilter(mf,Np=5000)))
  ll <- logmeanexp(evals,se=TRUE)
  mf |> coef() |> bind_rows() |>
    bind_cols(loglik=ll[1],loglik.se=ll[2])
} -> results

#store in our dataframe
read_csv(csv_name, show_col_types = FALSE) |>
  bind_rows(results) |>
  arrange(-loglik) |>
  write_csv(csv_name)


Output[[model]][["results_loc"]] <- results
print(max(results$loglik))
```



```{r}
lik_form <- "~loglik"
for (p in names(coef(runsPOMP[[model]]))) {
  lik_form <- paste0(lik_form, "+", p)
}
lik_form <- as.formula(lik_form)

pairs(lik_form,data=results, pch=16)
```


```{r}
set.seed(657)
guesses <- global_guesses(model)
mf1 <- mifs_local[[1]]
```

```{r}
foreach(guess=iterators::iter(guesses,"row"), .combine=rbind,
    .options.future=list(seed=1312)
) %dofuture% {
    mf1 |>
      mif2(params=c(guess)) |>
      mif2(Nmif=100) -> mf
    replicate(
      10,
      mf |> pfilter(Np=5000) |> logLik()
  ) |>
    logmeanexp(se=TRUE) -> ll
  mf |> coef() |> bind_rows() |>
    bind_cols(loglik=ll[1],loglik.se=ll[2])
} -> results

Output[[model]][["results_glob"]] <- results
```



```{r}
print(max(results$loglik))
```



```{r}
results |>
  filter(loglik>max(loglik)-50) |>
  bind_rows(guesses) |>
  mutate(type=if_else(is.na(loglik),"guess","result")) |>
  arrange(type) -> all

Output[[model]][["all"]] <- all

pairs(lik_form, data=all, pch=16, cex=0.3,
      col=ifelse(all$type=="guess",grey(0.5),"red"))
```


```{r}
saveRDS(Output[[model]], paste0("Output_", model, ".RDS"))
```

